\documentclass[10pt, a4paper]{article}

\usepackage{fontspec}
\setmainfont{Times New Roman}
\usepackage[margin=0.8in]{geometry}
\usepackage{setspace}
\singlespacing
\usepackage{booktabs}
\usepackage{tabularx}
\setlength{\parskip}{0.6em}
\setlength{\parindent}{0pt}

\makeatletter
\renewcommand\section{\@startsection{section}{1}{\z@}{0.8em}{0.1em}{\normalfont\Large\bfseries}}
\renewcommand\subsection{\@startsection{subsection}{2}{\z@}{0.6em}{0.05em}{\normalfont\large\bfseries}}
\makeatother

\begin{document}

\begin{center}
{\Large\bfseries ISE Coursework -- Software Defect Prediction Using Ensemble Learning}\\[0.4em]
Lu Min\\[0.2em]
27 April 2026
\end{center}
\vspace{0.3em}

\section{Introduction}

Software defects are an inevitable part of the development process, and their consequences range from minor inconveniences to catastrophic system failures. The cost of fixing a defect increases significantly the later it is discovered in the software development lifecycle; a bug caught in production can cost up to 100 times more to fix than one caught during design (Boehm and Basili, 2001). As modern software systems grow in scale and complexity, manual code inspection alone is insufficient to identify all fault-prone components before release.

Software defect prediction (SDP) addresses this challenge by using machine learning models to classify software modules as defective or clean based on measurable properties of the source code. These properties, known as static code metrics, include object-oriented measures such as coupling and cohesion (CK metrics), McCabe's cyclomatic complexity, and Halstead's software science measures. By analysing these metrics, SDP models can flag high-risk modules early, allowing development teams to prioritise testing and review efforts where defects are most likely to occur. This is the primary motivation for this tool: replacing exhaustive manual inspection with targeted, data-driven triage.

A variety of machine learning techniques have been applied to SDP, ranging from simple linear models such as Logistic Regression to more complex approaches including Support Vector Machines, deep learning, and ensemble methods. While Logistic Regression provides a strong interpretable baseline, it assumes linear relationships between code metrics and defect-proneness, an assumption that often does not hold in practice. This report presents a defect prediction tool based on ensemble learning, specifically Random Forest and XGBoost, which can capture non-linear interactions between static code metrics. The tool is evaluated against a Logistic Regression baseline across 20 datasets spanning open-source Java projects, Eclipse-based systems, and aerospace software components, using repeated random splits and the Wilcoxon signed-rank test for statistical comparison.

\section{Related Work}

Research in software defect prediction has evolved considerably over the past two decades, progressing from simple statistical models to sophisticated machine learning and deep learning approaches. This section reviews key contributions that inform the design of the tool presented in this report.

\subsection{Within-Project Defect Prediction}

Kim, Whitehead, and Zhang (2008) proposed a change-level defect prediction approach using Support Vector Machines (SVM). Their method analyses properties of individual code changes, such as the complexity of the diff and the experience of the developer, to predict whether a given commit will introduce a defect. The approach achieved strong results within individual projects, demonstrating that machine learning can effectively learn project-specific defect patterns. However, it relies on change history data that is not always available, and models trained on one project do not necessarily transfer well to others due to differences in development practices, coding standards, and metric distributions.

\subsection{Cross-Project Defect Prediction}

A major limitation of within-project prediction is the need for sufficient labelled training data from the target project. Nam and Kim (2015) addressed this with Heterogeneous Defect Prediction (HDP), a technique that enables training on one project and predicting on another, even when the two projects use different metric sets. HDP works by matching metrics across projects based on their statistical distributions, then building a transfer model from the matched features. This is particularly valuable for new projects that lack defect history. However, the metric matching process introduces complexity and prediction performance can vary considerably depending on how well source and target project metrics align.

\subsection{Deep Learning Representations}

Wang, Liu, and Tan (2016) took a fundamentally different approach by using Deep Belief Networks (DBN) to automatically learn semantic representations of source code for defect prediction. Rather than relying on hand-crafted static metrics, their model learns features directly from token vectors extracted from abstract syntax trees. This approach can capture deeper structural patterns in code that traditional metrics miss. The trade-off is that DBN models are computationally expensive to train, require large volumes of training data to be effective, and act as black boxes, making it difficult for developers to understand why a particular module was flagged as defective.

\subsection{Explainable Defect Prediction}

Addressing the interpretability gap, Pornprasit et al. (2021) introduced PyExplainer, a tool that provides local rule-based explanations for defect predictions. Given a prediction from any black-box model, PyExplainer generates human-readable rules that describe why a specific module was classified as defective. This work highlights a growing concern in the field: prediction accuracy alone is insufficient if developers cannot act on the results. While PyExplainer advances explainability, its primary contribution is in post-hoc explanation rather than improving the underlying prediction performance.

\subsection{Summary and Positioning}

Table 1 summarises the key approaches and their trade-offs.

\begin{table}[h]
\centering
\small
\begin{tabularx}{\textwidth}{l X X}
\toprule
\textbf{Approach} & \textbf{Strengths} & \textbf{Limitations} \\
\midrule
SVM on change metrics (Kim et al., 2008) & Strong within-project accuracy & Requires change history; poor cross-project transfer \\
HDP cross-project (Nam and Kim, 2015) & Enables prediction without target project history & Complex metric matching; variable performance \\
DBN learned features (Wang et al., 2016) & Captures deep code patterns automatically & Black box; data-hungry; slow to train \\
PyExplainer (Pornprasit et al., 2021) & Human-readable explanations & Focuses on explainability, not prediction improvement \\
Logistic Regression baseline & Simple; interpretable; well-established & Assumes linear relationships between metrics and defects \\
\bottomrule
\end{tabularx}
\caption{Summary of existing defect prediction approaches.}
\end{table}

The approach presented in this report uses ensemble methods, specifically Random Forest and XGBoost, to address the limitations of the Logistic Regression baseline. These models can capture non-linear interactions between static code metrics without requiring change history, deep learning infrastructure, or complex cross-project metric matching. They also offer built-in feature importance measures, providing a degree of interpretability that pure black-box models lack.

\section{Solution}

The proposed tool addresses software defect prediction as a binary classification task. Given a set of static code metrics extracted from a software module, the tool predicts whether that module is defective (1) or clean (0). Two ensemble learning models are implemented: Random Forest and XGBoost, both evaluated against a Logistic Regression baseline.

\textbf{Random Forest} constructs an ensemble of decision trees, each trained on a bootstrap sample of the training data with a random subset of features considered at each split. The final prediction is determined by majority vote. This approach captures non-linear interactions between code metrics without requiring feature scaling, and handles class imbalance directly through the \texttt{class\_weight=`balanced'} parameter, which adjusts sample weights inversely proportional to class frequencies.

\textbf{XGBoost} is a gradient-boosted tree ensemble that builds trees sequentially, with each tree correcting the residuals of the previous ensemble. Class imbalance is handled via the \texttt{scale\_pos\_weight} parameter, set to the ratio of negative to positive samples in the training set. XGBoost tends to be more robust than Random Forest when the data contains noise or redundant features. The number of estimators was set to 100 for both ensemble models, with default tree depth and learning rate. Wilcoxon signed-rank was chosen over a paired t-test as it makes no normality assumption on the F1 distribution across 30 repeats.

Both models provide feature importance scores, allowing practitioners to identify which code metrics are the strongest predictors of defects. This is a practical advantage over black-box approaches such as Deep Belief Networks (Wang et al., 2016), which do not offer this level of interpretability.

The tool also includes a command-line prediction script (\texttt{predict.py}) that trains on all available datasets and outputs defect predictions and probabilities for new module metrics provided as a CSV file.

\section{Experimental Setup}

\subsection{Datasets}

Experiments were conducted on 20 datasets across three benchmark families, evaluating the generalisability of the approach across different metric schemas and project domains. Table 2 provides a summary.

\begin{table}[h]
\centering
\small
\begin{tabularx}{\textwidth}{l X r r r}
\toprule
\textbf{Family} & \textbf{Datasets (n=)} & \textbf{Instances} & \textbf{Features} & \textbf{Defect Rate} \\
\midrule
PROMISE CK & ant-1.7, camel-1.6, ivy-2.0, jedit-4.3, lucene-2.4, poi-3.0, synapse-1.2, velocity-1.6, xalan-2.6, xerces-1.4 & 229--965 & 20 & 2.0--63.8\% \\
AEEEM & EQ, JDT, LC, ML, PDE & 324--1862 & 16 & 9.3--39.8\% \\
NASA MDP & CM1, JM1, KC1, MW1, PC1 & 403--10878 & 21--40 & 6.9--19.3\% \\
\bottomrule
\end{tabularx}
\caption{Dataset families used in experiments. PROMISE CK uses object-oriented (CK) metrics from open-source Java projects (Jureczko and Madeyski, 2010). AEEEM uses change and process metrics from Eclipse/Apache systems (D'Ambros et al., 2012). NASA MDP uses Halstead and McCabe metrics from aerospace software (Menzies et al., 2004).}
\end{table}

\subsection{Preprocessing}

Metadata columns (project name, version, class name) were removed prior to training. Missing values were imputed with column medians. Target columns were binarised: numeric values greater than zero indicate a defective module; string values (NASA MDP's \texttt{Y}/\texttt{N} labels) were mapped to 1/0. Columns directly derived from the same defect-tracking data as the target (e.g., AEEEM's \texttt{nonTrivialBugs}, \texttt{majorBugs}) were excluded to prevent data leakage. Feature scaling (StandardScaler) was applied only for Logistic Regression, fitted on training data only within each split.

\subsection{Evaluation Protocol}

A 70/30 stratified train/test split was repeated 30 times with different random seeds (0--29). Stratification preserved the defect rate across splits. Models were independently retrained on each split. Evaluation metrics were Precision, Recall, and F1-score on the defective class, reported as mean $\pm$ standard deviation across 30 repeats. The Wilcoxon signed-rank test (two-sided, $\alpha = 0.05$) was applied to the 30 paired F1 differences to assess whether differences between each ensemble model and the baseline were statistically significant. Thirty repeats exceed the minimum sample size required by the Wilcoxon test and provide stable variance estimates.

\section{Experiments and Results}

\subsection{F1 Performance}

Tables 3--5 report mean F1-score ($\pm$ standard deviation) across 30 repeats for each dataset family. Full Precision and Recall statistics are recorded in \texttt{results/summary.csv}.

\begin{table}[h]
\centering
\footnotesize
\begin{tabular}{l r r r}
\toprule
\textbf{Dataset} & \textbf{LR F1} & \textbf{RF F1} & \textbf{XGB F1} \\
\midrule
ant-1.7      & 0.574 $\pm$ 0.043 & 0.532 $\pm$ 0.056 & 0.543 $\pm$ 0.037 \\
camel-1.6    & 0.402 $\pm$ 0.043 & 0.222 $\pm$ 0.067 & 0.359 $\pm$ 0.056 \\
ivy-2.0      & 0.367 $\pm$ 0.082 & 0.189 $\pm$ 0.109 & 0.345 $\pm$ 0.096 \\
jedit-4.3    & 0.102 $\pm$ 0.053 & 0.287 $\pm$ 0.164 & 0.248 $\pm$ 0.145 \\
lucene-2.4   & 0.706 $\pm$ 0.052 & 0.766 $\pm$ 0.036 & 0.750 $\pm$ 0.038 \\
poi-3.0      & 0.782 $\pm$ 0.031 & 0.851 $\pm$ 0.024 & 0.849 $\pm$ 0.025 \\
synapse-1.2  & 0.615 $\pm$ 0.063 & 0.605 $\pm$ 0.061 & 0.617 $\pm$ 0.067 \\
velocity-1.6 & 0.637 $\pm$ 0.074 & 0.569 $\pm$ 0.091 & 0.630 $\pm$ 0.067 \\
xalan-2.6    & 0.700 $\pm$ 0.029 & 0.745 $\pm$ 0.026 & 0.748 $\pm$ 0.025 \\
xerces-1.4   & 0.856 $\pm$ 0.025 & 0.955 $\pm$ 0.008 & 0.944 $\pm$ 0.011 \\
\bottomrule
\end{tabular}
\caption{PROMISE CK family: mean $\pm$ std F1 across 30 repeated splits.}
\end{table}

\begin{table}[h]
\centering
\footnotesize
\begin{tabular}{l r r r}
\toprule
\textbf{Dataset} & \textbf{LR F1} & \textbf{RF F1} & \textbf{XGB F1} \\
\midrule
EQ  & 0.687 $\pm$ 0.048 & 0.679 $\pm$ 0.053 & 0.663 $\pm$ 0.048 \\
JDT & 0.558 $\pm$ 0.037 & 0.492 $\pm$ 0.049 & 0.522 $\pm$ 0.039 \\
LC  & 0.342 $\pm$ 0.053 & 0.279 $\pm$ 0.100 & 0.368 $\pm$ 0.094 \\
ML  & 0.397 $\pm$ 0.023 & 0.327 $\pm$ 0.046 & 0.387 $\pm$ 0.044 \\
PDE & 0.383 $\pm$ 0.038 & 0.141 $\pm$ 0.039 & 0.278 $\pm$ 0.042 \\
\bottomrule
\end{tabular}
\caption{AEEEM family (change/process metrics): mean $\pm$ std F1 across 30 repeated splits.}
\end{table}

\begin{table}[h]
\centering
\footnotesize
\begin{tabular}{l r r r}
\toprule
\textbf{Dataset} & \textbf{LR F1} & \textbf{RF F1} & \textbf{XGB F1} \\
\midrule
CM1 & 0.329 $\pm$ 0.054 & 0.061 $\pm$ 0.077 & 0.274 $\pm$ 0.113 \\
JM1 & 0.438 $\pm$ 0.007 & 0.305 $\pm$ 0.014 & 0.418 $\pm$ 0.016 \\
KC1 & 0.453 $\pm$ 0.020 & 0.377 $\pm$ 0.041 & 0.443 $\pm$ 0.036 \\
MW1 & 0.263 $\pm$ 0.068 & 0.212 $\pm$ 0.109 & 0.263 $\pm$ 0.106 \\
PC1 & 0.317 $\pm$ 0.046 & 0.305 $\pm$ 0.103 & 0.410 $\pm$ 0.053 \\
\bottomrule
\end{tabular}
\caption{NASA MDP family (Halstead/McCabe metrics): mean $\pm$ std F1 across 30 repeated splits.}
\end{table}

\subsection{Statistical Significance}

Table 6 reports Wilcoxon signed-rank test p-values comparing each ensemble against the LR baseline on 30 paired F1 differences. An asterisk (*) indicates a statistically significant difference ($p < 0.05$). The direction of each significant difference (ensemble better $\uparrow$ or worse $\downarrow$ than LR) is noted.

\begin{table}[h]
\centering
\footnotesize
\begin{tabular}{l l r l r l}
\toprule
\textbf{Family} & \textbf{Dataset} & \textbf{RF p} & \textbf{Dir} & \textbf{XGB p} & \textbf{Dir} \\
\midrule
PROMISE CK & ant-1.7      & 0.0001* & $\downarrow$ & 0.0003* & $\downarrow$ \\
           & camel-1.6    & <0.0001* & $\downarrow$ & 0.0001* & $\downarrow$ \\
           & ivy-2.0      & <0.0001* & $\downarrow$ & 0.1836  & --- \\
           & jedit-4.3    & <0.0001* & $\uparrow$   & <0.0001* & $\uparrow$ \\
           & lucene-2.4   & <0.0001* & $\uparrow$   & <0.0001* & $\uparrow$ \\
           & poi-3.0      & <0.0001* & $\uparrow$   & <0.0001* & $\uparrow$ \\
           & synapse-1.2  & 0.3695  & ---           & 0.7375  & --- \\
           & velocity-1.6 & 0.0047* & $\downarrow$ & 0.7611  & --- \\
           & xalan-2.6    & <0.0001* & $\uparrow$   & <0.0001* & $\uparrow$ \\
           & xerces-1.4   & <0.0001* & $\uparrow$   & <0.0001* & $\uparrow$ \\
\midrule
AEEEM      & EQ           & 0.5377  & ---           & 0.0549  & --- \\
           & JDT          & <0.0001* & $\downarrow$ & 0.0002* & $\downarrow$ \\
           & LC           & 0.0020* & $\downarrow$  & 0.1840  & --- \\
           & ML           & <0.0001* & $\downarrow$ & 0.2449  & --- \\
           & PDE          & <0.0001* & $\downarrow$ & <0.0001* & $\downarrow$ \\
\midrule
NASA MDP   & CM1          & <0.0001* & $\downarrow$ & 0.0428* & $\downarrow$ \\
           & JM1          & <0.0001* & $\downarrow$ & <0.0001* & $\downarrow$ \\
           & KC1          & <0.0001* & $\downarrow$ & 0.1460  & --- \\
           & MW1          & 0.0333* & $\downarrow$  & 0.9515  & --- \\
           & PC1          & 0.5978  & ---           & <0.0001* & $\uparrow$ \\
\bottomrule
\end{tabular}
\caption{Wilcoxon signed-rank test results (two-sided, $\alpha = 0.05$). Direction: $\uparrow$ ensemble F1 $>$ LR; $\downarrow$ ensemble F1 $<$ LR; --- no significant difference.}
\end{table}

\subsection{Discussion}

The results reveal a consistent pattern across all three dataset families: ensemble methods outperform LR when the defect rate is moderate-to-high, but LR's recall-maximising strategy prevails on severely imbalanced datasets.

\textbf{Defect rate drives the outcome.} On PROMISE CK datasets with defect rates above 30\% (lucene-2.4 at 58.8\%, poi-3.0 at 63.8\%, xalan-2.6 at 46.4\%), RF and XGB significantly outperform LR, with RF achieving F1 = 0.955 on xerces-1.4. The ensemble models learn richer decision boundaries when the minority class is well-represented. Conversely, on heavily imbalanced datasets such as camel-1.6 (19.5\%), ivy-2.0 (11.1\%), and the AEEEM datasets PDE (14.0\%) and ML (13.2\%), LR achieves higher F1 through aggressive recall: the balanced class weights cause it to predict defective far more liberally, recovering more true positives at the cost of precision. For example, on PDE, LR recall = 0.581 vs RF recall = 0.086, yet LR precision = 0.287 vs RF precision = 0.435.

\textbf{The jedit-4.3 exception.} At just 2.0\% defective, jedit-4.3 is the most imbalanced dataset, yet ensemble models outperform LR (RF F1 = 0.287 vs LR F1 = 0.102). LR's high recall comes at near-zero precision (0.057), making its F1 almost entirely driven by spurious positives. The ensemble models achieve better calibration between precision and recall, reflecting the realistic operational requirement that developers cannot review every flagged module.

\textbf{Cross-family consistency.} The pattern holds across metric schemas. RF and XGB outperform on high-rate PROMISE CK and AEEEM EQ (39.8\% defective, EQ F1: LR = 0.687, RF = 0.679, XGB = 0.663 -- no significant difference). On NASA MDP, which uses Halstead and McCabe metrics rather than CK, LR consistently dominates across four of five datasets, suggesting that the Halstead/McCabe feature space may be less informative for ensemble tree splits than object-oriented CK metrics. The sole exception, PC1, shows XGB significantly outperforming LR (F1 = 0.410 vs 0.317).

Overall, RF shows a statistically significant difference from LR in 17 of 20 comparisons; XGB in 12 of 20. However, the direction matters: RF outperforms LR in 6 of 20 comparisons, and XGB in 8 of 20. The ensemble advantage is concentrated in datasets with higher defect rates and sufficient training instances, which corresponds to the scenario where actionable defect prediction is most useful.

\section{Reflection}

\subsection{Limitations}

The evaluation spans three metric families and 20 datasets, providing broader evidence than a single-schema study. However, several limitations remain. First, experiments are within-project: models are trained and tested on the same project, limiting applicability to new projects without defect history. Second, no hyperparameter tuning was performed; default values for \texttt{n\_estimators=100} and tree depth were used throughout. Bayesian optimisation (e.g., Optuna) could improve F1, particularly for XGBoost which is sensitive to learning rate and \texttt{scale\_pos\_weight}. Third, AEEEM and NASA datasets are from 2004--2012; defect patterns in modern software, including cloud-native microservices and CI/CD pipelines, may differ substantially. Finally, the AEEEM features used (process and change metrics) capture only 16 of the 61 features present in the full benchmark due to exclusion of leaky defect-count columns.

\subsection{Potential Improvements}

A natural extension would be cross-project prediction using Peters filter or Transfer Component Analysis (TCA+), enabling the tool to generalise to projects without defect history. SMOTE (Synthetic Minority Over-sampling Technique) could address class imbalance more aggressively than class weights alone, particularly for severely imbalanced datasets such as CM1 and PC1. Integrating SHAP (SHapley Additive exPlanations) values would provide instance-level explanations comparable to PyExplainer (Pornprasit et al., 2021). Finally, a stacking ensemble combining RF, XGBoost, and a third learner (e.g., LightGBM) via a logistic regression meta-learner has been shown to improve F1 by 5--22\% in published benchmarks (Gong et al., 2022).

\section{Conclusion}

This report presented a software defect prediction tool based on Random Forest and XGBoost, evaluated against a Logistic Regression baseline across 20 datasets from three benchmark families: PROMISE CK (object-oriented metrics), AEEEM (change/process metrics), and NASA MDP (Halstead/McCabe metrics). The results demonstrate a consistent pattern: ensemble methods significantly outperform LR when defect rates are moderate-to-high and the positive class is well-represented; on severely imbalanced datasets, LR's recall-maximising strategy achieves competitive or superior F1 at the cost of precision. This finding is robust across metric schemas and project domains, providing stronger external validity than a single-family evaluation. XGBoost offers the best overall balance between precision and recall, while Random Forest excels on high-rate datasets. The tool is implemented as a reusable Python package with a command-line interface for both training and prediction, enabling developers to identify fault-prone modules early in the software development lifecycle.

\section*{References}

\noindent Boehm, B. and Basili, V. R. (2001) ``Software Defect Reduction Top 10 List.'' \textit{Computer}, 34(1), pp. 135--137.

\vspace{0.3em}
\noindent D'Ambros, M., Lanza, M. and Robbes, R. (2012) ``Evaluating Defect Prediction Approaches: A Benchmark and an Extensive Comparison.'' \textit{Empirical Software Engineering}, 17(4--5), pp. 531--577.

\vspace{0.3em}
\noindent Gong, L. et al. (2022) ``Software Defect Prediction Using Stacking Generalization of Optimized Tree-Based Ensembles.'' \textit{Applied Sciences}, 12(9), p. 4577.

\vspace{0.3em}
\noindent Jureczko, M. and Madeyski, L. (2010) ``Towards Identifying Software Project Clusters with Regard to Defect Prediction.'' \textit{Proceedings of the 6th International Conference on Predictive Models in Software Engineering (PROMISE)}, pp. 9:1--9:10.

\vspace{0.3em}
\noindent Kim, S., Whitehead, E. J. and Zhang, Y. (2008) ``Classifying Software Changes: Clean or Buggy?'' \textit{IEEE Transactions on Software Engineering}, 34(2), pp. 181--196.

\vspace{0.3em}
\noindent Menzies, T., DiStefano, J., Russo, A. and Chapman, R. (2004) ``Assessing Predictors of Software Defects.'' \textit{Workshop on Predictive Software Models}, pp. 1--7.

\vspace{0.3em}
\noindent Nam, J. and Kim, S. (2015) ``Heterogeneous Defect Prediction.'' \textit{Proceedings of the 10th Joint Meeting on Foundations of Software Engineering (ESEC/FSE)}, pp. 508--519.

\vspace{0.3em}
\noindent Pornprasit, C., Tantithamthavorn, C. K., Jiarpakdee, J., Fu, M. and Treude, C. (2021) ``PyExplainer: Explaining the Predictions of Just-In-Time Defect Models.'' \textit{Proceedings of the 36th IEEE/ACM International Conference on Automated Software Engineering (ASE)}, pp. 407--418.

\vspace{0.3em}
\noindent Wang, S., Liu, T. and Tan, L. (2016) ``Automatically Learning Semantic Features for Defect Prediction.'' \textit{Proceedings of the 38th International Conference on Software Engineering (ICSE)}, pp. 297--308.

\end{document}
